---
title: "Intermediate Elective"
author: "Zelda Reif"
date: "2026-05-22"
format: pdf
---

# Set Up

## Packages

```{r}
#| label: packages

# general use
library(tidyverse)

# file organization
library(here)

# reading .xlsx files 
library(readxl)

# data visualization
library(naniar)
library(patchwork)
library(NatParksPalettes)
library(gghighlight)
```

## Data

```{r}
#| label: data

# vegetation
veg <- read_csv(here("data", "veg copy.csv"))

# veg metadata
metadata <- read_csv(here("data", "vp_veg_metadata.csv"))

# taxonomic info
taxon_list <- read_csv(here("data","taxon_list.csv"))

# aquatic invertebrates
aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Aquatic Insects")
```

# Problems

## 1. Explain your inspiration

I was inspired by Steven Ponce’s “Evolution of MTA Art Materials (1980-2020)” because the figure's characteristics would allow me to show total percent cover of multiple vernal pool vegetation species through time, while highlighting one species per faceted plot.

This figure makes sense for the vernal pool vegetation data set because it allows for a clear comparison of total percent cover between different species with distinct lines.

The plot will be faceted by vegetation species, the x axis will be time by year and the y axis will be total percent cover. Each plotted line will represent a different vernal pool vegetation species, and the species of interest's line will be highlighted with a different color to stand out. 

## 2. 

Sketch what your figure would look like following your inspiration.

Insert that sketch (as a photo, screenshot, or otherwise) into your document.



## 3. Code your figure

```{r}
#| label: wrangling-veg

# create new object from veg
native_spp <- veg |>
  # filter to only include native cover
  filter(str_detect(cover_category, "NATIVE COVER")) |>
  # filter by species
  pull(psoc) |>
  # list each unique native species
  unique()

# display 10 random species from data frame
sample(native_spp, 10)



# create new object from veg
native_pc <- veg |>
# filter to only include native species
filter(psoc %in% native_spp) |>
# filtered to only include ncos vernal pool transects
filter(str_detect(transect_name, "VP") &
site == "ncos") |>
group_by(transect_name, year) |>
# summarized by calculating total percent cover across all
# quadrats
summarise(sum_pc = sum(percent_cover, na.rm = TRUE)) |>
# un-grouped
ungroup() |>
# created column to join with metadata
unite("year_pool", year, transect_name, remove = TRUE) |>
# joined metadata data frame
left_join(metadata, by = "year_pool") |>
# calculated mean percent cover
mutate(mean_pc = sum_pc/num_quad)




```


```{r}
#| label: plotting-veg-pc
#| fig-height: 9
#| fig-width: 9

  title_text <- "Total Native Percent Cover Over Time"
subtitle_text <- "NCOS vernal pool transects"

caption_text <- "Vegetation survey data"


### |- plot ----


  ggplot(data = veg_pc,
    mapping = aes(x = year, 
                  y = mean_pc)) +

  # Geoms
  geom_line(linewidth = 1.3, color = "#4A9782", alpha = 0.9) +
  geom_point(color = "#4A9782", size = 2, alpha = 0.8) +

  # Scales
  scale_x_continuous(expand = expansion(mult = c(0.01, 0.01))) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +

  # relabeling x- and y-axes, adding title
  labs(x = "Year",
      y = "Total Percent Cover",
      title = "Total Native Species Percent Cover at NCOS 2021-25") +


  # Theme (clean + stable)
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 11),
    plot.caption = element_text(size = 8),
    axis.title = element_text(face = "bold")
  )

p
```

## 4. Describe your process

In 1-3 sentences each, describe:

- your process of using someone else’s code to create a visualization
- challenges you encountered as you made your visualization
- how your visualization differs from visualizations you made in class, for assignments, or for a previous elective