---
title: "Intermediate Elective"
format: pdf
---

# Set Up

## Packages

```{r}
#| label: packages

# general use
library(tidyverse)

# file organization
library(here)

# reading .xlsx files 
library(readxl)

# data visualization
library(naniar)
library(patchwork)
```

## Data

```{r}
#| label: data

# vegetation
veg <- read_csv(here("data", "veg copy.csv"))

# veg metadata
metadata <- read_csv(here("data", "vp_veg_metadata.csv"))

# taxonomic info
taxon_list <- read_csv(here("data","taxon_list.csv"))

# aquatic invertebrates
aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Aquatic Insects")
```

# Problems

## 1. Explain your inspiration

I was inspired by Steven Ponce’s “Evolution of MTA Art Materials (1980-2020)” because the figure's characteristics would allow me to show multiple metrics (average abundance/L for multiple invertebrate species) for each location at NCOS.

This figure makes sense for the aquatic invertebrate data set because it allows for the separation of data by specific location at NCOS while still allowing for clear comparison of abundance between different species with distinct lines.

The plot will be faceted by location, the x axis will be time by year and the y axis will be average abundance/L. Each plotted line will represent a different invertebrate species, and will be a different color. The species corresponding to each color will be indicated by a legend.

## 2. 

Sketch what your figure would look like following your inspiration.

Insert that sketch (as a photo, screenshot, or otherwise) into your document.



## 3. Code your figure

## 4. Describe your process

In 1-3 sentences each, describe:

- your process of using someone else’s code to create a visualization
- challenges you encountered as you made your visualization
- how your visualization differs from visualizations you made in class, for assignments, or for a previous elective