---
title: "Intermediate Elective"
author: "Zelda Reif"
date: "2026-05-22"
format: pdf
---

# Set Up

## Packages

```{r}
#| label: packages

# general use
library(tidyverse)

# file organization
library(here)

# reading .xlsx files 
library(readxl)

# data visualization
library(naniar)
library(patchwork)
```

## Data

```{r}
#| label: data

# vegetation
veg <- read_csv(here("data", "veg copy.csv"))

# veg metadata
metadata <- read_csv(here("data", "vp_veg_metadata.csv"))

# taxonomic info
taxon_list <- read_csv(here("data","taxon_list.csv"))

# aquatic invertebrates
aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Aquatic Insects")
```

# Problems

## 1. Explain your inspiration

I was inspired by Steven Ponce’s “Evolution of MTA Art Materials (1980-2020)” because the figure's characteristics would allow me to show total percent cover of multiple vernal pool vegetation species through time, while highlighting one species per faceted plot.

This figure makes sense for the vernal pool vegetation data set because it allows for a clear comparison of total percent cover between different species with distinct lines.

The plot will be faceted by vegetation species, the x axis will be time by year and the y axis will be total percent cover. Each plotted line will represent a different vernal pool vegetation species, and the species of interest's line will be highlighted with a different color to stand out. 

## 2. 

Sketch what your figure would look like following your inspiration.

Insert that sketch (as a photo, screenshot, or otherwise) into your document.



## 3. Code your figure

```{r}

```


## 4. Describe your process

In 1-3 sentences each, describe:

- your process of using someone else’s code to create a visualization
- challenges you encountered as you made your visualization
- how your visualization differs from visualizations you made in class, for assignments, or for a previous elective