---
title: "Intermediate Elective"
author: "Zelda Reif"
date: "2026-05-22"
format: pdf
---

# Set Up

## Packages

```{r}
#| label: packages

# general use
library(tidyverse)

# file organization
library(here)

# reading .xlsx files 
library(readxl)

# data visualization
library(naniar)
library(patchwork)
```

## Data

```{r}
#| label: data

# vegetation
veg <- read_csv(here("data", "veg copy.csv"))

# veg metadata
metadata <- read_csv(here("data", "vp_veg_metadata.csv"))

# taxonomic info
taxon_list <- read_csv(here("data","taxon_list.csv"))

# aquatic invertebrates
aq_ins <- read_xlsx(
  here("data", "Aquatic Sampling Data-2026-03-10.xlsx"),
  sheet = "Aquatic Insects")
```

# Problems

## 1. Explain your inspiration

I was inspired by Tobias Stalder’s “Hiking Locations in Washington” because the figure's characteristics would allow me to show average abundance/L for an invertebrate species for each location at NCOS.

This figure makes sense for the aquatic invertebrate data set because it allows for the separation of data by specific location at NCOS while still allowing for clear comparison of abundance between different species with distinct differences in visual weight throughout the plot.

For the plot, the x axis will be the location and the y axis will be average abundance/L. The species corresponding to each color will be indicated by a legend.

## 2. 

Sketch what your figure would look like following your inspiration.

Insert that sketch (as a photo, screenshot, or otherwise) into your document.



## 3. Code your figure

```{r}

```


## 4. Describe your process

In 1-3 sentences each, describe:

- your process of using someone else’s code to create a visualization
- challenges you encountered as you made your visualization
- how your visualization differs from visualizations you made in class, for assignments, or for a previous elective